PAMELA: an annotation-based Java Modeling Framework

### 2021-03-23 6-Experimentations.tex
review up to p13 (but the quotation mark)

The full Openflexo infrastructure is composed of about 50 components. In most components, the PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes. 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.

@@

An interesting experimentation has been done in the context of Openflexo development process. When PAMELA has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

>>>>

The full Openflexo infrastructure is composed of about 50 components. In most components, the PAMELA framework is extensively used. As an example the \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes. 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.

@@

An interesting experiment has been done in the context of Openflexo development process. When PAMELA was integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

### 2021-03-23 Todo-review2.tex
review up to p13 (but the quotation mark)

There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:
- "to /avoid" on p3 (DONE)
- "detailed thereafter" -> "detailed hereafter" (It's Correct)
- "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
- "with setter/getter": make both plural (DONE)
- POJO used here and elsewhere yet the acronym is introduced much later (DONE)
- p4 l11-12: reduce etc. should be reduce_s_ etc. (Keep as it is)
- odd spacing before : consistently throughout bullet list + text  (Keep as it is: I trust Latex)
- "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela -> PAMELA ; to be checked)
- "defines API": insert "an" (DONE)
- "in Java interface": insert "a" (DONE)
- "returning desired value", "taking value" etc.: insert "the" (DONE)
- "May also declares"... (DONE)
- p6 l46: @Embedded/"as embedded" meaning?
- p7 "of launched Java application" missing "the"
- "is/are" -> "is (are)", in line with parentheses earlier in sentence
- "Experimented programmers": I doubt it, unless you got major ethical clearance
- "Listing 2: model" -> "~: Model" consistent with Listing 1
- the book title is "Lord of the ring_s_" with s
- "using classical Java ... scheme": missing "the"
- "Java programming language": insert "The"
- l43 inconsistent spacing
- l46 package ... package
- p11 l36 sticking into margin
- S.4.3 please add size in LoC and classes
- java -> Java
- p13 "diana component": insert "the"
- "experimentation" -> "experiment"
- "When Pamela has been" -> "After Pamela was"
- l36 the quotation mark before code generation is the wrong one
- p 14 "appertaining" -> "pertaining", "relating"

>>>>

There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:
- "to /avoid" on p3 (DONE)
- "detailed thereafter" -> "detailed hereafter" (It's Correct : As adverbs the difference between hereafter and thereafter is that hereafter is in time to come; in some future time or state while thereafter is after that, from then on.)
- "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
- "with setter/getter": make both plural (DONE)
- POJO used here and elsewhere yet the acronym is introduced much later (DONE)
- p4 l11-12: reduce etc. should be reduce_s_ etc. (Keep as it is)
- odd spacing before : consistently throughout bullet list + text  (Keep as it is: I trust Latex)
- "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela -> PAMELA ; to be checked)
- "defines API": insert "an" (DONE)
- "in Java interface": insert "a" (DONE)
- "returning desired value", "taking value" etc.: insert "the" (DONE)
- "May also declares"... (DONE)
- p6 l46: @Embedded/"as embedded" meaning?
- p7 "of launched Java application" missing "the" (DONE)
- "is/are" -> "is (are)", in line with parentheses earlier in sentence (DONE)
- "Experimented programmers": I doubt it, unless you got major ethical clearance (Replaced by "Experienced PAMELA programmers")
- "Listing 2: model" -> "~: Model" consistent with Listing 1 (DONE)
- the book title is "Lord of the ring_s_" with s (DONE)
- "using classical Java ... scheme": missing "the" (Check :mis au pluriel)
- "Java programming language": insert "The" (DONE)
- l43 inconsistent spacing (DONE)
- l46 package ... package (DONE)
- p11 l36 sticking into margin (Final check to do)
- S.4.3 please add size in LoC and classes (?)
- java -> Java (DONE)
- p13 "diana component": insert "the" (DONE)
- "experimentation" -> "experiment" (DONE)
- "When Pamela has been" -> "After Pamela was" (DONE : if we consider it's over)
- l36 the quotation mark before code generation is the wrong one
- p 14 "appertaining" -> "pertaining", "relating"

### 2021-03-23 architecture.tex
review up to p13 (but the quotation mark)

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May also declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, eventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique re-indexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
      embedded. %according to PAMELA semantics.
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.

@@

The PAMELA model at runtime is computed dynamically, working on the classpath of
launched Java application, and starting from a simple Java interface (or a
collection of Java interfaces) which is/are PAMELA-annotated. From a
mathematical point of view, internal representation of the underlying model is
a graph whose vertices are PAMELA \emph{ModelEntities} (annotated Java
interface), and edges are either inheritance links or reference links (a
property whose type is another \emph{ModelEntity}). \mytexttt{@Imports} and
  \mytexttt{@Import} annotations allow to include some other
  \emph{ModelEntities} in the model. An annotation attribute
  \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that
  context, PAMELA model computation is a graph closure computation, starting
  from a collection of vertices.

@@

Programmers may already reuse a number of advanced programming features
(some already mentioned) such as: multiple inheritance and traits, containment
management, cloning, fine-grained notification management, object graph
comparison and diff/merge, visiting patterns, clipboard management, validation,
support for \emph{design by contract} by integrating assertions from Java
Modeling Language (JML), Metaprogramming, Aspect Oriented Programming. Each of
those features is described on a dedicated page that is reachable from the
official web site\footnote{\url{https://pamela.openflexo.org}}.
Experimented programmers may extend PAMELA defining new annotations.

@@

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=model execution/manipulation, label=lst:execution]
// Instantiate the meta-model
// by computing the closure of concepts graph
ModelContext modelContext 
    = ModelContextLibrary.getModelContext(Library.class);
// Instantiate the factory
ModelFactory factory  = new ModelFactory(modelContext);
// Instantiate a Library
Library myLibrary = factory.newInstance(Library.class);
// Instantiate some Books
Book myFirstBook 
    = factory.newInstance(Book.class, "Lord of the ring");
Book anOtherBook = factory.newInstance(Book.class, "Holy bible");
myLibrary.addToBooks(myFirstBook);
myLibrary.addToBooks(anOtherBook);
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

@@

Custom code can be easily added to this model as we show in Listing~\ref{lst:custom}. It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity. Custom implementations are defined using classical Java implementation/overrides scheme. Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, cannot be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references a default interpreted implementation (call to \texttt{performSuperGetter(String)} method).

>>>>

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May also declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, eventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique re-indexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
      embedded \todo{meaning?}. %according to PAMELA semantics.
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.

@@

The PAMELA model at runtime is computed dynamically, working on the classpath of
the launched Java application, and starting from a simple Java interface (or a
collection of Java interfaces) which is (are) PAMELA-annotated. From a
mathematical point of view, internal representation of the underlying model is
a graph whose vertices are PAMELA \emph{ModelEntities} (annotated Java
interface), and edges are either inheritance links or reference links (a
property whose type is another \emph{ModelEntity}). \mytexttt{@Imports} and
  \mytexttt{@Import} annotations allow to include some other
  \emph{ModelEntities} in the model. An annotation attribute
  \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that
  context, PAMELA model computation is a graph closure computation, starting
  from a collection of vertices.

@@

Programmers may already reuse a number of advanced programming features
(some already mentioned) such as: multiple inheritance and traits, containment
management, cloning, fine-grained notification management, object graph
comparison and diff/merge, visiting patterns, clipboard management, validation,
support for \emph{design by contract} by integrating assertions from Java
Modeling Language (JML), Metaprogramming, Aspect Oriented Programming. Each of
those features is described on a dedicated page that is reachable from the
official web site\footnote{\url{https://pamela.openflexo.org}}.
Experienced PAMELA programmers may extend PAMELA defining new annotations.

@@

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=Model execution/manipulation, label=lst:execution]
// Instantiate the meta-model
// by computing the closure of concepts graph
ModelContext modelContext 
    = ModelContextLibrary.getModelContext(Library.class);
// Instantiate the factory
ModelFactory factory  = new ModelFactory(modelContext);
// Instantiate a Library
Library myLibrary = factory.newInstance(Library.class);
// Instantiate some Books
Book myFirstBook 
    = factory.newInstance(Book.class, "Lord of the rings");
Book anOtherBook = factory.newInstance(Book.class, "Holy bible");
myLibrary.addToBooks(myFirstBook);
myLibrary.addToBooks(anOtherBook);
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

@@

Custom code can be easily added to this model as we show in Listing~\ref{lst:custom}. It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity. Custom implementations are defined using classical Java implementation/override schemes. Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, cannot be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references a default interpreted implementation (call to \texttt{performSuperGetter(String)} method).

### 2021-03-23 implementation.tex
review up to p13 (but the quotation mark)

The model-code integration we advocate requires facilities to encode metadata in the source code. This requires an annotation-enabled language. Such a language supports the attribute-oriented programming if its grammar allows adding custom declarative tags to annotate standard program elements. Java programming language is a good candidate, as it supports annotations.

The PAMELA API exposed to the developer mainly consists of : 1) a set of annotations; and 2) a set of unimplemented Java interfaces exposing required features.

The package \texttt{org.openflexo.pamela.annotations} package exposes the set of annotations which were presented in Subsection \ref{sub:DesignTime}.

The package \texttt{org.openflexo.pamela} contains the following feature-related Java interfaces:
\begin{itemize}
    \item \texttt{AccessibleProxyObject} is the interface that PAMELA objects should extend in order to benefit from base features such as generic default implementation, containment management, notification, object graph comparison and diff/merge, visiting patterns, etc.
    \item \texttt{CloneableProxyObject} exposes features related to cloning.
    \item \texttt{DeletableProxyObject} exposes features related to deletion management.
    \item \texttt{SpecifiableProxyObject} exposes dynamic assertion checking features in the context of JML (contract management) use. 
\end{itemize}

@@

Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).

>>>>

The model-code integration we advocate requires facilities to encode metadata in the source code. This requires an annotation-enabled language. Such a language supports the attribute-oriented programming if its grammar allows adding custom declarative tags to annotate standard program elements. The Java programming language is a good candidate, as it supports annotations.

The PAMELA API exposed to the developer mainly consists of: 1) a set of annotations; and 2) a set of unimplemented Java interfaces exposing required features.

The \texttt{org.openflexo.pamela.annotations} package exposes the set of annotations which were presented in Subsection \ref{sub:DesignTime}.

The \texttt{org.openflexo.pamela} package contains the following feature-related Java interfaces:
\begin{itemize}
    \item \texttt{AccessibleProxyObject} is the interface that PAMELA objects should extend in order to benefit from base features such as generic default implementation, containment management, notification, object graph comparison and diff/merge, visiting patterns, etc.
    \item \texttt{CloneableProxyObject} exposes features related to cloning.
    \item \texttt{DeletableProxyObject} exposes features related to deletion management.
    \item \texttt{SpecifiableProxyObject} exposes dynamic assertion checking features in the context of JML (contract management) use. 
\end{itemize}

@@

Compared to a basic POJO implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two Java implementations have been derived using code generation. The first implementation was made using fully implemented Java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
